Accelerated Inference for Partially Observed Stochastic Processes using Automatic Differentiation

### 2024-06-11 paper/appendix/targeting.tex
a few typos, a minor title change, a note in appendix D

\begin{defn}[Targeting]
    A random vector $(X, w)$ drawn from a distribution $g$ \textbf{targets} the distribution $\pi$ if for any measurable and bounded functional $h$,
\begin{equation}
    E_g\{h(X) \cdot w\}=E_\pi\{h(X)\}
\end{equation}  
    A set of particles $(X^J_j, w^J_j), j=1,2, \ldots,J$, \textbf{targets} $\pi$ if for any measurable and bounded functional $h$,
\begin{equation}
    \frac{\sum_{j=1}^J h(X^J_j) w^J_j}{\sum_{j=1}^J w^J_j} \stackrel{a.s.}{\to} E_\pi(h(X))
\end{equation}
as $J \to \infty$.
\end{defn}
\cite{chopin04} asserted without proof that common particle filter algorithms target the filtering distribution, $f_{X_{1:N}|Y_{1:N}}$, in this sense.
In order to prove the consistency of our variation on the particle filter, we now present three helper lemmas.
The first follows from standard importance sampling arguments, the second from integrating out the marginal, and the third from Bayes' theorem. 
We state Lemma~\ref{lem:change-measure-proper-weights} assuming multinomial resampling, which is convenient for the proof though other resampling strategies may be preferable in practice.

@@

\begin{lem}[Particle Marginals]
    \label{lem:marginal-proper-weights}
    Suppose that $\{(\tilde X_j^J,U_j^J),j=1,\dots,J\}$ targets $f_X$. Also suppose that $\tilde Z_j^J \sim f_{Z|X}(\cdot | \tilde X_j^J)$ where $f_{Z|X}$ is a conditional probability density function corresponding to a joint density $f_{X,Z}$ with marginal densities $f_X$ and $f_Z$. Then, if the $U_j^J$ are bounded, $\{(\tilde Z_j^J,U_j^J)\}$ targets $f_Z$.
\end{lem}
\begin{proof}
    We want to show that, for any measurable bounded $h$, 
     $$\frac{\sum_j h(\tilde Z_j^J) \, U_j^J}{\sum_j U_j^J} \stackrel{a.s.}{\to} \E_{f_Z}[h(Z)] = \E_{f_X}\big[\E_{f_{Z|X}}[h(Z) | X]\big].$$
    By assumption, for any measurable and bounded functional $g$ with domain $\gX$,
    \begin{equation}\label{eq:lemma2:g}
    \frac{\sum_j g(\tilde X_j^J)\, U_j^J}{\sum_j U_j^J} \stackrel{a.s.}{\to} \E_{f_X}[g(X)].
    \end{equation}
    Let $\bar{U}_j^J = \frac{J U_j^J}{\sum_j U_j^J}$. Examine the numerator and denominator of the quantity $$\frac{J^{-1}\sum_j h(\tilde Z_j^J) \, U_j^J}{J^{-1}\sum_j U_j^J} = \frac{J^{-1}\sum_j h(\tilde Z_j^J) \, \bar{U}_j^J}{J^{-1}\sum_j \bar{U}_j^J}.$$
    The denominator converges to $1$ almost surely. The numerator, on the other hand, is
    $$\frac{1}{J}\sum_j h(\tilde Z_j^J)\, \bar{U}_j^J,$$
    by the same fourth moment argument to the above lemma converges almost surely to the limit of its expectation
    \begin{eqnarray*}        
    \lim_{J\to\infty} \E\left[\frac{1}{J}\sum_j h(\tilde Z_j^J)\, \bar{U}_j^J\right] 
    &=& \lim_{J\to\infty} \E \left[\frac{1}{J}\sum_j  
      \E\left[ h(\tilde Z_j^J)\, \bar{U}_j^J\big|\tilde X_j^J, \bar U_j^J\right]
    \right]
    \\
    &=& \lim_{J\to\infty} \E \left[\frac{1}{J}\sum_j  \E\big[h(Z) \big| X=\tilde X_j^J \big] \, \bar U_j^J\right].
    \end{eqnarray*}
    Applying (\ref{eq:lemma2:g}) with $g(x) = \E\big[h(Z) | X=x\big]$, the average on the right hand size converges almost surely to $\E\big\{\E[h(Z)|X]\big\}=\E[h(Z)].$
    It remains to swap the limit and expectations. We can do so with the bounded convergence theorem, and therefore obtain   
    $$\frac{1}{J}\sum_j h(\tilde Z_j^J)\,\bar{U}_j^J \stackrel{a.s.}{\to} \E_{f_Z}[h(Z)].$$

>>>>

\begin{defn}[Targeting]
    A random vector $(X, w)$ drawn from a distribution $g$ \textbf{targets} the distribution $\pi$ if for any measurable and bounded functional $h$,
\begin{equation}
    E_g\{h(X) \cdot w\}=E_\pi\{h(X)\}
\end{equation}  
    A set of particles $(X^J_j, w^J_j), j=1,2, \ldots,J$, \textbf{targets} $\pi$ if for any measurable and bounded functional $h$,
\begin{equation}
    \frac{\sum_{j=1}^J h(X^J_j) w^J_j}{\sum_{j=1}^J w^J_j} \stackrel{a.s.}{\to} E_\pi(h(X))
\end{equation}
as $J \to \infty$.
\end{defn}
\cite{chopin04} \ed{and \cite{liu01}????} asserted without proof that common particle filter algorithms target the filtering distribution, $f_{X_{1:N}|Y_{1:N}}$, in this sense.
\ed{\cite{chopin20} proved a related result assuming bounded densities.
  We follow a similar approach to \cite{chopin20}, based on a strong laws of large numbers for triangular arrays.
  Triangular array strong laws do not hold without an additional regularity condition such as boundedness.}
\ed{I PROPOSE ADDING SOMETHING LIKE THIS, PARTLY TO MOTIVATE THE RESULTS, PARTLY TO BE KINDER TO CHOPIN.}  
In order to prove the consistency of our variation on the particle filter, we now present three helper lemmas.
The first follows from standard importance sampling arguments, the second from integrating out the marginal, and the third from Bayes' theorem. 
We state Lemma~\ref{lem:change-measure-proper-weights} assuming multinomial resampling, which is convenient for the proof though other resampling strategies may be preferable in practice.

@@

\begin{lem}[Particle Marginals]
    \label{lem:marginal-proper-weights}
    Suppose that $\{(\tilde X_j^J,U_j^J),j=1,\dots,J\}$ targets $f_X$. Also suppose that $\tilde Z_j^J \sim f_{Z|X}(\cdot | \tilde X_j^J)$ where $f_{Z|X}$ is a conditional probability density function corresponding to a joint density $f_{X,Z}$ with marginal densities $f_X$ and $f_Z$. Then, if the $U_j^J$ are bounded, $\{(\tilde Z_j^J,U_j^J)\}$ targets $f_Z$.
\end{lem}
\begin{proof}
    We want to show that, for any measurable bounded $h$, 
     $$\frac{\sum_j h(\tilde Z_j^J) \, U_j^J}{\sum_j U_j^J} \stackrel{a.s.}{\to} \E_{f_Z}[h(Z)] = \E_{f_X}\big[\E_{f_{Z|X}}[h(Z) | X]\big].$$
    By assumption, for any measurable and bounded functional $g$ with domain $\gX$,
    \begin{equation}\label{eq:lemma2:g}
    \frac{\sum_j g(\tilde X_j^J)\, U_j^J}{\sum_j U_j^J} \stackrel{a.s.}{\to} \E_{f_X}[g(X)].
    \end{equation}
    Let $\bar{U}_j^J = \frac{J U_j^J}{\sum_j U_j^J}$. Examine the numerator and denominator of the quantity $$\frac{J^{-1}\sum_j h(\tilde Z_j^J) \, U_j^J}{J^{-1}\sum_j U_j^J} = \frac{J^{-1}\sum_j h(\tilde Z_j^J) \, \bar{U}_j^J}{J^{-1}\sum_j \bar{U}_j^J}.$$
    The denominator converges to $1$ almost surely. The numerator, on the other hand, is
    $$\frac{1}{J}\sum_j h(\tilde Z_j^J)\, \bar{U}_j^J,$$
    by the same fourth moment argument to the above lemma converges almost surely to the limit of its expectation
    \begin{eqnarray*}        
    \lim_{J\to\infty} \E\left[\frac{1}{J}\sum_j h(\tilde Z_j^J)\, \bar{U}_j^J\right] 
    &=& \lim_{J\to\infty} \E \left[\frac{1}{J}\sum_j  
      \E\left[ h(\tilde Z_j^J)\, \bar{U}_j^J\big|\tilde X_j^J, \bar U_j^J\right]
    \right]
    \\
    &=& \lim_{J\to\infty} \E \left[\frac{1}{J}\sum_j  \E\big[h(Z) \big| X=\tilde X_j^J \big] \, \bar U_j^J\right].
    \end{eqnarray*}
    Applying (\ref{eq:lemma2:g}) with $g(x) = \E\big[h(Z) | X=x\big]$, the average on the right hand side converges almost surely to $\E\big\{\E[h(Z)|X]\big\}=\E[h(Z)].$
    It remains to swap the limit and expectations. We can do so with the bounded convergence theorem, and therefore obtain   
    $$\frac{1}{J}\sum_j h(\tilde Z_j^J)\,\bar{U}_j^J \stackrel{a.s.}{\to} \E_{f_Z}[h(Z)].$$

### 2024-06-11 paper/ms.tex
a few typos, a minor title change, a note in appendix D

\title{Accelerated Inference for Partially Observed Stochastic Processes with Automatic Differentiation}

@@

Scibior and Wood \cite{scibior21} showed that the estimators derived by Poyiadjis et al. \cite{poyiadjis11} can be attained with standard AD software, using an algorithmic prodecure called a ``stop-gradient'' which allows selected exressions to be evaluated but not differentiated.
Unfortunately, \cite{scibior21} used their estimator within an algorithm which does not have the plug-and-play property and which has high Monte Carlo variability.
We use \cite{scibior21} and \cite{poyiadjis11} as our starting point, while developing a new approach which remedies the weaknesses of these papers.

@@

\begin{table}[h!]
\centering
\input{imgs/095/table}
\caption{Maximum log-likelihood found by IF2, IFAD, and MOP alone. IFAD performs the best among all methods. Our implementation of IF2 outperforms that of \cite{ionides15}, but still ultimately underperforms IFAD. IFAD manages to find the MLE, matching the highest log-likelihood previously found in the \texttt{dacca()} model implemented within the \texttt{pomp} package of \cite{king16}.}
\label{table:mle}
\end{table}

@@

\begin{figure}[ht]
    \centering
    \includegraphics[width=\arxiv{10cm}{\textwidth/3}]{imgs/095/optim.png}
    \caption{Optimization progress of IFAD and IF2. The dashed orange line depicts the median warm-start given by running 40 IF2 iterations. While running 60 more iterations of IF2 improves upon the median warm-start, doing so ultimately underperforms IFAD. We see that IFAD has better tail control and successfully reaches the MLE. 
    We use a dotted red line to display the MLE.}
    \label{fig:optim}
\end{figure}

>>>>

\title{Accelerated Inference for Partially Observed Stochastic Processes using Automatic Differentiation}

@@

Scibior and Wood \cite{scibior21} showed that the estimators derived by Poyiadjis et al. \cite{poyiadjis11} can be attained with standard AD software, using an algorithmic procedure called a ``stop-gradient'' which allows selected exressions to be evaluated but not differentiated.
Unfortunately, \cite{scibior21} used their estimator within an algorithm which does not have the plug-and-play property and which has high Monte Carlo variability.
We use \cite{scibior21} and \cite{poyiadjis11} as our starting point, while developing a new approach which remedies the weaknesses of these papers.

@@

\begin{table}[h!]
\centering
\input{../imgs/095/table}
\caption{Maximum log-likelihood found by IF2, IFAD, and MOP alone. IFAD performs the best among all methods. Our implementation of IF2 outperforms that of \cite{ionides15}, but still ultimately underperforms IFAD. IFAD manages to find the MLE, matching the highest log-likelihood previously found in the \texttt{dacca()} model implemented within the \texttt{pomp} package of \cite{king16}.}
\label{table:mle}
\end{table}

@@

\begin{figure}[ht]
    \centering
    \includegraphics[width=\arxiv{10cm}{\textwidth/3}]{../imgs/095/optim.png}
    \caption{Optimization progress of IFAD and IF2. The dashed orange line depicts the median warm-start given by running 40 IF2 iterations. While running 60 more iterations of IF2 improves upon the median warm-start, doing so ultimately underperforms IFAD. We see that IFAD has better tail control and successfully reaches the MLE. 
    We use a dotted red line to display the MLE.}
    \label{fig:optim}
\end{figure}
